---
title: "Admissions Bias Analysis at UC Berkeley"
author: "Antonia the First of Surrey"
format: 
  html:
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=Great+Vibes&display=swap" rel="stylesheet">
        <style>
          h1.title {
            font-family: 'Cinzel', serif;
            font-weight: 700;
            font-size: 2.5em;
          }
        </style>
editor: visual
---

# **Introduction**

We've been tasked by the court to analyze UC Berkeley's admissions data in response to a 1973 lawsuit claiming sex-based discrimination. The plaintiff asserts that female applicants face a disadvantage in the admissions process. The objective is clear: determine whether gender plays a significant role in admission outcomes at the university.

To provide a transparent analysis that can be reviewed by others, I will be using a Quarto notebook to document my code and findings.

```{r}
#| label: load-packages
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Data Overview

To begin, we load the admissions data provided by the court. This dataset includes information about applicants, their gender, and admission outcomes, allowing us to explore the relationship between gender and admission rates.

```{r}
#| label: load-data
url <- "https://waf.cs.illinois.edu/discovery/berkeley.csv"
data <- read.csv(url)
head(data)
summary(data)
```

## Admission Rates by Gender

We start by calculating the admission rates for each gender to explore potential disparities.

```{r}
#| label: admission-rates-gender
admission_rates_gender <- data %>%
  group_by(Gender) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )

ggplot(admission_rates_gender, aes(x = Gender, y = Admission_Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Gender", 
       x = "Gender", 
       y = "Admission Rate") +
  theme_minimal()
```

The bar chart above shows the admission rates for male and female applicants. We observe that the admission rate for male applicants is slightly higher than that for female applicants. Specifically, while both genders show considerable numbers of accepted applicants, males have a notably higher proportion admitted.

This difference in admission rates raises questions about potential gender-based disparities in the admissions process. While the visualization does not confirm discrimination, it suggests a trend where gender might influence admission outcomes. Further analysis is required, particularly to examine if this difference persists across different majors or if other factors might explain the variation.

## Admission Rates by Major

Now, we calculate the admission rates by major to identify if certain programs exhibit gender biases.

```{r}
#| label: admission-rates-major
admission_rate_major <- data %>%
  select(Major, Admission) %>%
  group_by(Major) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )

ggplot(admission_rate_major, aes(x = Major, y = Admission_Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Major", 
       x = "Major", 
       y = "Admission Rate") +
  theme_minimal()
```

The bar chart shows that admission rates vary widely across majors. Majors A and B have the highest admission rates, both above 60%, while Major F has the lowest. This variation suggests that some programs may have more selective or lenient admissions criteria, potentially impacting the overall fairness of the admissions process.

## Applications by Gender and Major

Here, we examine the total applications by major and gender to understand the gender distribution across programs.

```{r}
#| label: applications-gender-major
#| warning: false
applications_gender_major <- data %>%
  select(Major, Gender) %>%
  group_by(Major, Gender) %>%
  summarize(Total = n())

ggplot(applications_gender_major, aes(x = Major, y = Total, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Applications by Major and Gender", 
       x = "Major", 
       y = "Applications") +
  theme_minimal()
```

Here’s a discussion of results for the **Total Applications by Major and Gender** plot:

------------------------------------------------------------------------

This bar chart shows the distribution of applications by gender across each major. We can see that male applicants tend to have higher application numbers in most majors, particularly in "Other," where both male and female applications peak. Majors A and C also show gender disparities, with males applying more frequently than females.

These trends suggest possible gender preferences or imbalances in major selection, which could impact admission patterns and the overall gender distribution within certain fields. Understanding the reasons behind these gender-based preferences might be valuable for addressing potential imbalances.

## Admission Rates by Major and Gender

We further break down admission rates by major and gender to analyze any major-specific biases.

```{r}
#| label: admission-rates-gender-major
admission_rates_gender_major <- data %>%
  group_by(Major, Gender) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )

ggplot(admission_rates_gender_major, aes(x = Major, y = Admission_Rate, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Major and Gender", 
       x = "Major", 
       y = "Admission Rate") +
  theme_minimal()
```

The chart shows that admission rates vary by both major and gender. In Majors A and B, females have slightly higher admission rates than males, while other majors display smaller gender differences or consistently low rates (e.g., Major F). This suggests that some departments may have admission patterns that differ by gender, potentially contributing to overall gender disparities.

## Gender Difference in Admission Rates by Major

Finally, we examine the differences in admission rates by major between male and female applicants.

```{r}
#| label: gender-diff-major
#| warning: false
admission_rate_diff <- admission_rates_gender_major %>%
  select(Major, Gender, Admission_Rate) %>%
  pivot_wider(names_from = Gender, values_from = Admission_Rate) %>%
  mutate(Difference = M - F)

ggplot(admission_rate_diff, aes(x = Major, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("dark green", "skyblue"), 
                    labels = c("Female Higher", "Male Higher")) +  
  labs(title = "Difference Between Male and Female Admission Rates by Major",
       x = "Major", 
       y = "Difference (Male - Female)") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black", size = 1) +
  theme_minimal() +
  theme(legend.title = element_blank())
```

The chart shows the difference in admission rates between males and females across majors. Majors A and B have higher admission rates for females, as indicated by the negative values, while Majors C, E, and "Other" show higher rates for males. This pattern suggests that gender-based differences in admissions vary by department, which could influence overall gender representation within these fields.

# Conclusion

The analysis reveals notable gender-based patterns in UC Berkeley's admissions process. While some majors, like A and B, show higher admission rates for females, others, such as C and E, favor male applicants. Additionally, overall admission rates vary significantly by major, suggesting that both departmental criteria and gender may influence admission outcomes. These findings highlight areas where further investigation into admission policies could help ensure a more equitable process.
